---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bruneimap

<!-- badges: start -->
[![R-CMD-check](https://github.com/propertypricebn/bruneimap/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/propertypricebn/bruneimap/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Provides Brunei GIS data in the form of a `sf` (simple features) object ready for plotting and analysis in R. 
The smallest (areal) unit of analysis is a "kampong", which in turn is contained within mukims and districts.

## Installation

You can install the development version of bruneimap from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pkg_install("propertypricebn/bruneimap")
```

## Features

There are three `sf` files contained in the package:

1. `dis_sf` (District level boundaries)
2. `mkm_sf` (Mukim level boundaries)
3. `kpg_sf` (Kampong level boundaries)

Most likely you will want to use the kampong level data. Of course, the `kpg_sf` can be summarised in terms of mukims and districts, but for convenience the other two data sets are provided as well. 

### Data

```{r, message = FALSE}
# Load libraries
library(tidyverse)
theme_set(theme_bw())
library(bruneimap)
library(sf)

# What's in our data set?
glimpse(kpg_sf)
```

### Plots

```{r brunei_map_plot, message = FALSE, warning = FALSE}
ggplot(kpg_sf) +
  geom_sf(aes(fill = mukim)) +
  geom_sf(data = mkm_sf, col = "black", lwd = 0.5, fill = NA) +
  theme(legend.position = "none") +
  scale_fill_viridis_d(option = "turbo")
```


## Acknowledgements

- The Brunei map was generated from publically available [Survey Department](https://geoportal.survey.gov.bn/start-gp) data. The GeoJSON files were sourced from [thewheat/brunei_map](https://github.com/thewheat/brunei_map) GitHub repo. Comes with a neat [web app](http://thewheat.github.io/brunei_map/). Many thanks to these guys!
