\documentclass[12pt]{article}
\usepackage[a4paper, margin = 1in]{geometry}
\usepackage{graphicx, amsmath, booktabs}
\usepackage{biblatex} 


\title{Spatial data analysis of school (locations?) in Brunei Darussalam }
\author{Alvin Bong JL, Aniq Najwa, Danish Ikhwan, Rayme Hijazi}
\date{November 2024}

\setlength{\parindent}{0pt} % Remove paragraph indentation globally
\begin{document}

\maketitle
\begin{abstract}
This study analyzes the spatial distribution of schools across Brunei Darussalam to assess how well current educational facilities serve both urban and rural populations. Using spatial autocorrelation techniques like Moran’s I and the Getis-Ord \( G_i^* \) statistic, this research identifies clusters of schools and evaluates their alignment with population density. Findings indicate significant school clustering in urban areas, particularly near Bandar Seri Begawan, with comparatively fewer facilities in rural regions. This imbalance suggests potential challenges in accessing education for students in less populated areas. The study offers valuable insights for policymakers to improve resource distribution, aiming to support equitable access to education and inform future spatial planning in educational infrastructure. 
\end{abstract}

%keywords section
%cover page for report

\section{Introduction}
Education plays a critical role in fostering social and economic development, and equitable access to educational facilities is essential for ensuring equal opportunities for all students. In Brunei Darussalam, education is free for citizens and compulsory from ages 5 to 16. The country’s high literacy rate, robust educational system, and strong economic standing position it as a leader in Southeast Asian education. However, while Brunei’s achievements in educational access are widely recognized, little research has been conducted to determine whether the spatial distribution of schools is optimized to serve both urban and rural populations effectively. \\

This study seeks to explore the distribution of schools in Brunei through three main questions:
\begin{enumerate}
    \item Are there clusters of schools within certain areas of Brunei?
    \item Where are these clusters located?
    \item Do these clusters align with population density?
\end{enumerate}

Brunei’s administrative divisions include districts, mukims (sub-districts), and kampongs (villages), each with varied population densities and geographic characteristics. For example, population concentrations are generally higher along the coastlines, particularly in the capital, Bandar Seri Begawan, while inland areas tend to be more sparsely populated. Given Brunei’s unique geography, understanding how educational facilities are distributed across these areas can provide insights into the accessibility challenges faced by students in different regions. \\

In the study of GIS and spatial data analysis, Dr. Haziq Jamil (2024) highlights the importance of examining spatial patterns to address questions of access and resource allocation. This study builds on this approach by applying spatial autocorrelation methods to assess school clustering and distribution across Brunei. Although this research does not address school quality directly, it provides essential foundational data on spatial distribution that can support future investigations into educational outcomes and equity in Brunei. By evaluating the alignment between school locations and population density, this study offers valuable insights that can guide policymakers in optimizing educational infrastructure to ensure equitable access for all students. \\

\section{Literature Review}
1. Motivation behind schools
2. History of spatial autocorrelation
3. Weights Matrix



\section{Description of Study Area and Data}
\subsection{Description of Study Area}
Brunei Darussalam, commonly known as Brunei, is located on the northern coast of the island of Borneo in Southeast Asia. With an area of approximately 5,765 square kilometers, Brunei is bordered by the South China Sea to the north and surrounded by the Malaysian state of Sarawak. The nation’s territory is divided into two non-contiguous areas:  The larger western section comprising Brunei-Muara, Tutong, and Belait districts; and the smaller eastern Temburong district. In the Northeast of the larger section lies Brunei's capital, Bandar Seri Begawan. \\

The districts of Brunei are subdivided into 39 smaller administrative zones known as mukims, each embraces a number of kampung (villages). Brunei’s geography is characterized by a mix of urban centers, dense forests, and coastal lowlands. More than 70\% of the nation is covered with forests, with majority locating inland, southern parts of Belait and Temburong districts. \\

% pic
as delineated and shown in Figure 1. green areas show forest. \\

% Population
According to census 2021, the population of Brunei stands at 445, 000. The population is concentrated along the coastline, especially around Bandar Seri Begawan, which serves as the nation’s administrative, cultural, and economic center. \\

% Wealth 
By major accounts, Brunei may be considered a high-income country. The Sultanate boasts the second-highest per capita income and Human Development Index (HDI) in Southeast Asia and the highest per capita Gross National Income (GNI) among OECD countries between 2005 and 2020 [43]. \\
 
 %Education
With education that is both free (for Brunei citizens) and compulsory from age 5 to 16, the vast majority of Brunei’s population is literate. In addition to the government schools are private Chinese schools and schools operated by religious institutions; all, however, generally follow the same government-developed curriculum. \\

Given the high wealth and free and robust education, it would be interesting to leverage spatial analysis in finding patterns and understanding how schools are clustered and distributed across the country.\\

\subsection{Data Collection \& Processing}
The dataset comprises \( N = 252 \) schools in Brunei Darussalam, sourced from the Ministry of Education’s \textit{2018 Education Statistics Book}. The decision to use the 2018 dataset stems from the lack of detailed data in more recent publications, which only provide summary versions. Specifically, the 2018 dataset includes:
\begin{enumerate}
    \item A complete listing of all schools in Brunei,
    \item A breakdown of schools by type (Ministry of Education [MOE], Ministry of Religious Affairs [MORA], and private),
    \item Categorization of MOE pre-primary to secondary schools into administrative clusters (Cluster 1–6), and
    \item Student-teacher ratios by cluster and type.
\end{enumerate}
,details which are not available in later editions of the statistical book. \\

Since the \textit{2018 Education Statistics Book} is only available in PDF format, the data was converted to an .xlsx file using an online converter. The data was then reorganized in Excel before being imported into R using the \texttt{readxl()} function. \\

%on collecting locations?%
Although we attempted to retrieve the locations (latitude and longitude) of all schools in Brunei using the \texttt{osmdata\_sf()} function from the \textit{osmdata} package in R, this approach proved incomplete. Some schools were missing, others were inaccurately located, and certain school names were abbreviated. Consequently, we only obtained partial location data. To address this, we used a \texttt{left\_join()} to merge the available locations with the school listing, and manually collected the remaining coordinates. \\


\subsection{Preliminary Data Analysis}
other than 2 special schools, schools in Brunei are categorized into 3 main sectors, MOE, MORA, private. The numbers of school are 164, 9 , 77. respectively, 65\%MOE 30\%private. The preprimary to six formth public schools in brunei (MOE Sector) are  further dsitributed into cluster 1-6. As seen is fig.x there is some overlap/ number of schools are quite even per cluster/ no other data on top of enrollment, we cant do much analysis?
% but can do global morann, compare cluster per areas to enrolment? 
% not really cuz the moran i use is areal
\\ 

But limited analysis can be done since there is slight overlap and insufficient data. Only have enrolment by cluster but not teachers therefore the following student teacher ratio is by type and district. The actual paper focuses all schools in general (in section 5,6?) \\ 
%maybe can consider global moran by sectors: MOE vs private? only 2 ctg, waste time?

As student/teachers/structure of vocational schools and higher education varies (depending on classes/ not consistent), this section (Preliminary Data Anlaysis : student teacher ratio section) focuses only on preprimary to sixth forms. \\

In regards to student teacher ratio, Belait and Brunei Muara districts have relatively higher values (at about 10) compared to Temburong and Tutong districts (approximately 7.6). Viewing this from sector pov, MOE and MORA school shares similar values. On the other hand, private schools generally have higher student teacher ratio, with the exception of Temburong district. 

% pic

\begin{table}[h!]
\centering
\setlength{\tabcolsep}{15pt} % Adjusts horizontal padding
\begin{tabular}{lrrr}
\hline
\textbf{Sector} & \textbf{Student} & \textbf{Teacher} & \textbf{Student-Teacher Ratio} \\ 
\hline
MOE             & 53,890           & 6,574            & 8.20                           \\
MORA            & 5,483            & 670              & 8.18                           \\
Private         & 32,692           & 2,315            & 14.1                           \\
\hline
\end{tabular}
\caption{Sector-wise Student-Teacher Ratio}
\end{table}

\begin{table}[h!]
\centering
\setlength{\tabcolsep}{10pt}
\begin{tabular}{lrrr}
\hline
\textbf{District}   & \textbf{Student}  & \textbf{Teacher}  & \textbf{Student-Teacher Ratio} \\ 
\hline
Belait              & 12,955            & 1,239             & 10.50 \\
Brunei Muara        & 68,188            & 6,892             & 9.89 \\
Temburong           & 1,893             & 248               & 7.63 \\
Tutong              & 9,029             & 1,180             & 7.65 \\ 
\hline
\end{tabular}
\caption{District-wise Student-Teacher Ratio}
\end{table}



\section{Methods}
We are interested in correlation between values of a variable (in our case count of schools per region) among spatially located data. Why? Judging by eyes is clearly not accurate and therefore the invention of spatial autocorrelation statistics. There are generally two categories of spatial autocorrelation statistics – global and local statistics. The former are estimated for the entire region under study, while the latter are estimated for individual locations to measure the relationship between the observations in the nearby area. (https://onlinelibrary.wiley.com/doi/10.1111/risa.17663)
assessing whether similar values cluster together across an entire study area. whether high or low values tend to be near each other or are randomly distributed. The test gives a single statistic for the entire dataset, summarizing the global spatial pattern.
identify whether there are clusters of similar values or spatial outliers at specific locations in a dataset


\subsection{Global test for spatial autocorrelation: Global Moran's I}
To examine whether schools in Brunei exhibit a clustered, dispersed, or random spatial pattern, we apply the Global Moran’s I test using the \texttt{global\_moran\_test()} function from the \texttt{sfdep} package in R. This test is computed for each mukim (administrative division) in the study area, indexed by \( i, j = 1, 2, \dots, M \). The Moran’s I test statistic is defined as follows: 

\begin{equation}
I = \frac{N}{S_0} \frac{\sum_{i=1}^N \sum_{j=1}^N w_{ij} (X_i - \bar{X})(X_j - \bar{X})}{\sum_{i=1}^N (X_i - \bar{X})^2}
\end{equation}

where:
\begin{itemize}
    \item \( X_i \) is the value of the study variable (count of schools) in area \( i \),
    \item \( \bar{X} \) is the mean of the study variable,
    \item \( w_{ij} \) is the spatial weight between mukims \( i \) and \( j \), with the weights matrix \( W \) defining the spatial relationship among regions.
\end{itemize}

For simplicity, a contiguity-based “rook” criterion is used for the spatial weights, as discussed in Section 1. %\ref{sec:literature-review} (Literature Review).% 
\\

Moran’s I values are standardized, with values close to \(+1\) indicating positive spatial autocorrelation (i.e., clustering), where high or low values are near each other. Values close to \(-1\) indicate negative spatial autocorrelation (i.e., dispersion), where neighboring values differ significantly. Values near \(0\) suggest randomness, indicating an absence of spatial pattern. \\

To determine the significance of the Moran’s I statistic, we employ the Central Limit Theorem to calculate p-values based on a Z-score, allowing us to test the following hypotheses:
\begin{itemize}
    \item \( H_0: I = 0 \) (no spatial autocorrelation),
    \item \( H_1: I \neq 0 \) (presence of spatial autocorrelation).
\end{itemize}




\subsection{Local indicators of spatial autocorrelation (LISA): Getis-Ord \( G_i^* \)}
While a visual inspection suggests that certain \textit{kampungs} (villages) may have a higher concentration of schools, we aim to quantify this pattern. Whereas global spatial autocorrelation tests confirm whether clustering exists, we use the Getis-Ord \( G_i^* \) statistic to identify the specific areas where schools are concentrated. This statistic is computed using the \texttt{hotspot\_gistar} function from the \texttt{sfhotspot} package in R. \\

In our analysis, the study area is subdivided into \( n \) rectangular grids, indexed by \( i = 1, 2, \dots, n \). For each \( i \)-th grid cell, the \( G_i^* \) statistic is calculated as:

\begin{equation}
G_i^* = \frac{\sum_j w_{ij} X_j}{\sum_j X_j}
\end{equation}

where:
\begin{itemize}
    \item \( X_j \) is the value of the study variable (i.e., the count of schools) for grid cell \( j \),
    \item \( w_{ij} \) is the spatial weight between grid cells \( i \) and \( j \), with 1 indicating neighboring cells and 0 for non-neighboring cells.
\end{itemize}

The spatial weights used are based on the same contiguity “rook” criterion applied in the Global Moran’s I test, with one modification: the spatial weight \( w_{ii} \) are set to 1 rather than 0. This adjustment gives \( G_i^* \) a more localized perspective, which is valuable for identifying clusters centered directly on a point of interest rather than merely in its surrounding areas. \\

A statistically significant high \( G_i^* \) value indicates a “hotspot,” or a cluster of high values, whereas a low \( G_i^* \) value suggests a “coldspot,” or a cluster of low values. \\

To only highlight significant clusters, the output was filtered to include only values with Gistar > xxx and p > 0.01?. The output dataset is cropped to Brunei’s boundary using \texttt{st\_intersection} to refine the analysis. This method identifies school hotspots, areas where there are more schools than would be expected if they were distributed randomly. \\



\section{Results}
The following section is split into 3 subsections corresponding to hypotheses as introduced in Section Introduction.

\subsection{Are there cluster?}
The Global Moran's I analysis yielded I value of 0.457. The positive Moran’s I statistic suggests a positive autocorrelation in count of schools across mukims. Given the statistically significant results (low p-value of $4.542 \times 10^{-6} < 0.001$ ), there is sufficient evidence to reject the null hypothesis, which assumes no spatial autocorrelation in the distribution of schools. \\

This finding supports the presence of moderate to strong clustering tendency, implying that mukims with a similar number of schools, whether high or low, are geographically close to each other. This results of the Global Moran’s I test aligns with our preliminary observation of school clusters being more concentrated near Brunei’s coastal regions, particularly along the South China Sea, as per outlined in Section EDA. In the following subsection, we will further verify this pattern through a localized analysis of clustering. \\

\subsection{Locations of School Clusters}
Fig. x highlights areas with school clusters that are higher than expected. The primary concentration is located in central Bandar Seri Begawan, the capital district of Brunei. This is expected its status as the nation’s capital, urban and administrative center.

Other notable clusters outside the capital are located in:
\begin{itemize}
    \item \textbf{Temburong District}: Mukim Bangar
    \item \textbf{Tutong District}: Mukim Pekan Tutong, Mukim Telisai
    \item \textbf{Kuala Belait District}: Mukim Kuala Belait, Mukim Seria
\end{itemize}

This result confirms out visual analysis outlined in the EDA section,  that school clusters tend to concentrate near the shoreline. \\

The Getis-Ord analysis offers an advantage over density maps by pinpointing the exact areas of clustering within each mukim. For example, in Kampong Panaga, schools are concentrated on the eastern side of the village rather than being evenly distributed. This level of detail enables a more precise understanding of spatial clustering patterns. \\

By providing this pinpoints of school clusters, it is hoped that this can help in future housing/amenity/etc planning? \\

%pic%

\subsection{Comparison to Distribution of Population}

%pic1 pic2%
Although visually, schools and population hotspots seems to match. When comparing the top 10 kampongs with more schools and population in Table. x below, we see that only 3 kampongs: Kampong A, Kg. B, Kg. C are in both categories. This suggests that clusters of schools are not located in exactly the areas with most population, but perhaps nearby. 


\section{Conclusions}

\section{Acknowledgements}
 The authors expresses gratitude OpenStreetMap contributors for the spatial data used in this analysis, and Dr Haziq Jamil, instructor of the module SM2302, for his invaluable insights
 which enriched this study.
 
\section{Reference}


\end{document}

